The Journey from Wild to Textbook Data: A Case Study from the National Longitudinal Survey of Youth

### 2021-09-21 paper/paper.Rmd
fixed begin{document} not found issue
replace table 1 as missing value figure 1

---
title: "The Journey from Wild to Textbook Data: A Case Study from the National Longitudinal Survey of Youth"
authors:
  - name: Dewi Amaliah
    thanks: ""
    department: Department of Econometrics and Business Statistics
    affiliation: Monash University
    location: Clayton, VIC 3800
    email: dama0007@student.monash.edu
  - name: Dianne Cook
    department: Department of Econometrics and Business Statistics
    affiliation: Monash University
    location: Clayton, VIC 3800
    email: dicook@monash.edu
  - name: Emi Tanaka
    department: Department of Econometrics and Business Statistics
    affiliation: Monash University
    location: Clayton, VIC 3800
    email: emi.tanaka@monash.edu
  - name: Kate Hyde
    thanks: ""
    department: Department of Econometrics and Business Statistics
    affiliation: Monash University
    location: Clayton, VIC 3800
    email: ""
  - name: Nicholas Tierney
    thanks: ""
    department: ""
    affiliation: ""
    location: ""
    email: ""
abstract: > 
 The National Longitudinal Survey of Youth (NLSY79) is a prominent open data source that has been important for educational purposes and multidisciplinary research on longitudinal data. Subsets of this data can be found in numerous textbooks and research articles. However, the steps and decisions taken to get from the raw data to the textbook data is never clearly articulated. This article describes our journey when trying to re-create a textbook example data set from the original database, with the goal being to refresh the textbook data more regularly. Thus, this paper demonstrates the process -- extracting, tidying, cleaning, and exploring with documentation -- to make refreshed data available for education or research. Three new data sets, and the code to produce them are provided in an accompanying open source R package, called `yowie`. As a result of this process, some recommendations are also made for the NLSY79 curators for incorporating data quality checks, and providing more convenient samples of the data to potential users. 
bibliography: references.bib
biblio-style: unsrt
#preamble: >
header-includes:
  - \usepackage{tcolorbox}
  - \usepackage{fontawesome}
output: 
  #rticles::arxiv_article:
    #keep_tex: false
  bookdown::pdf_book:
    base_format: rticles::arxiv_article
date: "24/02/2021"
always_allow_html: true
keywords: "Data cleaning; Data tidying; Longitudinal data; NLSY79; Open data; Initial data analysis; Outlier detection; Robust linear regression"
---

@@

In this section, we outline the steps taken to the download the raw data from the database provided at \url{https://www.nlsinfo.org/content/cohorts/nlsy79/get-data} [@nlsy79]. Specifically, we only extract a subset of the variables in the database as shown in \ref{box:data}.

## Getting and Tidying the Data

@@

::: {.tcolorbox data-latex="[label=\"box:data\", title=Getting the data]" }
\faDatabase National Longitudinal Survey of Youth 1979

@@

The downloaded data set came in a csv (NLSY79.csv) and dat (NLSY79.dat) format. We only used the .dat format. It also came along with these files:

- NLSY79.NLSY79: Tagset of variables that can be uploaded to the website to recreate the data set.
- NLSY79.R: R script provided automatically by the database for reading the data into R and converting the variables' names and label into something more sensible. We utilized this code to do an initial data tidying. It produced two data set,  `categories_qnames` (the observations are stored in categorical/interval values) and `new_data_qnames` (the observations are stored in integer form).

@@

```{r untidy-data}
```

>>>>

---
title: "The Journey from Wild to Textbook Data: A Case Study from the National Longitudinal Survey of Youth"
authors:
  - name: Dewi Amaliah
    thanks: ""
    department: Department of Econometrics and Business Statistics
    affiliation: Monash University
    location: Clayton, VIC 3800
    email: dama0007@student.monash.edu
  - name: Dianne Cook
    department: Department of Econometrics and Business Statistics
    affiliation: Monash University
    location: Clayton, VIC 3800
    email: dicook@monash.edu
  - name: Emi Tanaka
    department: Department of Econometrics and Business Statistics
    affiliation: Monash University
    location: Clayton, VIC 3800
    email: emi.tanaka@monash.edu
  - name: Kate Hyde
    thanks: ""
    department: Department of Econometrics and Business Statistics
    affiliation: Monash University
    location: Clayton, VIC 3800
    email: ""
  - name: Nicholas Tierney
    thanks: ""
    department: ""
    affiliation: ""
    location: ""
    email: ""
abstract: > 
 The National Longitudinal Survey of Youth (NLSY79) is a prominent open data source that has been important for educational purposes and multidisciplinary research on longitudinal data. Subsets of this data can be found in numerous textbooks and research articles. However, the steps and decisions taken to get from the raw data to the textbook data is never clearly articulated. This article describes our journey when trying to re-create a textbook example data set from the original database, with the goal being to refresh the textbook data more regularly. Thus, this paper demonstrates the process -- extracting, tidying, cleaning, and exploring with documentation -- to make refreshed data available for education or research. Three new data sets, and the code to produce them are provided in an accompanying open source R package, called `yowie`. As a result of this process, some recommendations are also made for the NLSY79 curators for incorporating data quality checks, and providing more convenient samples of the data to potential users. 
bibliography: references.bib
biblio-style: unsrt
#preamble: >
header-includes:
  - \usepackage{tcolorbox}
  - \usepackage{fontawesome}
  - \usepackage{color, colortbl}
output: 
  #rticles::arxiv_article:
    #keep_tex: false
  bookdown::pdf_book:
    base_format: rticles::arxiv_article
date: "24/02/2021"
always_allow_html: true
keywords: "Data cleaning; Data tidying; Longitudinal data; NLSY79; Open data; Initial data analysis; Outlier detection; Robust linear regression"
---

@@

In this section, we outline the steps taken to the download the raw data from the database provided at \url{https://www.nlsinfo.org/content/cohorts/nlsy79/get-data} [@nlsy79]. Specifically, we only extract a subset of the variables in the database as described in Section~\ref{getdata}.

## Getting and Tidying the Data {#getdata}

@@

::: {.tcolorbox data-latex="[label=\"box:data\", title=Getting the data]" }
\faDatabase\ National Longitudinal Survey of Youth 1979

@@

The downloaded data set came with the following set of files:

- `NLSY79.csv`: csv format
- `NLSY79.dat`: dat format
- `NLSY79.NLSY79`: Tagset of variables that can be uploaded to the website to recreate the data set.
- `NLSY79.R`: R script provided automatically by the database for reading the data into R and converting the variables' names and label into something more sensible. We utilized this code to do an initial data tidying. It produced two data set,  `categories_qnames` (the observations are stored in categorical/interval values) and `new_data_qnames` (the observations are stored in integer form).

@@

```{r untidy-data, fig.cap="The missing value structure of the NLSY79 raw data."}
```

### 2021-09-21 paper/paper.tex
replace table 1 as missing value figure 1

\usepackage{tcolorbox}
\usepackage{fontawesome}
\usepackage{booktabs}
\usepackage{longtable}
\usepackage{array}
\usepackage{multirow}
\usepackage{wrapfig}
\usepackage{float}
\usepackage{colortbl}
\usepackage{pdflscape}
\usepackage{tabu}
\usepackage{threeparttable}
\usepackage{threeparttablex}
\usepackage[normalem]{ulem}
\usepackage{makecell}
\usepackage{xcolor}

@@

In this section, we outline the steps taken to the download the raw data from the database provided at \url{https://www.nlsinfo.org/content/cohorts/nlsy79/get-data} (The U.S. Bureau of Labor Statistics, n.d.). Specifically, we only extract a subset of the variables in the database as described in Section\textasciitilde{}\ref{sec:getdata}.

@@

The downloaded data set came in a csv (NLSY79.csv) and dat (NLSY79.dat) format. We only used the .dat format. It also came along with these files:

\begin{itemize}
\tightlist
\item
  NLSY79.NLSY79: Tagset of variables that can be uploaded to the website to recreate the data set.
\item
  NLSY79.R: R script provided automatically by the database for reading the data into R and converting the variables' names and label into something more sensible. We utilized this code to do an initial data tidying. It produced two data set, \texttt{categories\_qnames} (the observations are stored in categorical/interval values) and \texttt{new\_data\_qnames} (the observations are stored in integer form).
\end{itemize}

@@

\caption{\label{tab:untidy-data}The Untidy Form of the NLSY79 Raw Data}
\centering
\begin{tabular}[t]{r|r|r|r|r|r|r}
\hline
CASEID\_1979 & HRP1\_1979 & HRP2\_1979 & HRP3\_1979 & HRP4\_1979 & HRP5\_1979 & HRP1\_1980\\
\hline
1 & 328 & NA & NA & NA & NA & NA\\
\hline
2 & 385 & NA & NA & NA & NA & 457\\
\hline
3 & 365 & NA & 275 & NA & NA & 397\\
\hline
4 & NA & NA & NA & NA & NA & NA\\
\hline
5 & 310 & 375 & NA & NA & NA & 333\\
\hline
6 & NA & NA & NA & 250 & NA & 275\\
\hline
\end{tabular}
\end{table}

>>>>

\usepackage{tcolorbox}
\usepackage{fontawesome}
\usepackage{color, colortbl}
\usepackage{booktabs}
\usepackage{longtable}
\usepackage{array}
\usepackage{multirow}
\usepackage{wrapfig}
\usepackage{float}
\usepackage{colortbl}
\usepackage{pdflscape}
\usepackage{tabu}
\usepackage{threeparttable}
\usepackage{threeparttablex}
\usepackage[normalem]{ulem}
\usepackage{makecell}
\usepackage{xcolor}

@@

In this section, we outline the steps taken to the download the raw data from the database provided at \url{https://www.nlsinfo.org/content/cohorts/nlsy79/get-data} (The U.S. Bureau of Labor Statistics, n.d.). Specifically, we only extract a subset of the variables in the database as described in Section\textasciitilde{}\ref{getdata}.

@@

The downloaded data set came with the following set of files:

\begin{itemize}
\tightlist
\item
  \texttt{NLSY79.csv}: csv format
\item
  \texttt{NLSY79.dat}: dat format
\item
  \texttt{NLSY79.NLSY79}: Tagset of variables that can be uploaded to the website to recreate the data set.
\item
  \texttt{NLSY79.R}: R script provided automatically by the database for reading the data into R and converting the variables' names and label into something more sensible. We utilized this code to do an initial data tidying. It produced two data set, \texttt{categories\_qnames} (the observations are stored in categorical/interval values) and \texttt{new\_data\_qnames} (the observations are stored in integer form).
\end{itemize}

@@

{\centering \includegraphics[width=468px]{figures/untidy-data-1}

\caption{The missing value structure of the NLSY79 raw data.}\label{fig:untidy-data}
\end{figure}
